---
title: "Wisconsin Cancer Data Analysis Project with DataCamp"
output: html_notebook
---

This is a guided project analyzing breast cancer data collected from the Wisconsin Breast Cancer dataset as hosted by Kaggle. Features are derived from digitized images from a fine needle aspirate describing the cell nuclei appearance. This project is an exercise in unsupervised machine learning, clustering, and PCA.

```{r}
url <- "https://assets.datacamp.com/production/course_1903/datasets/WisconsinCancer.csv"

# Download the data
wisc.df <- read.csv(url)

# Convert the features of the data
wisc.data <- as.matrix(wisc.df[3:32])

# Set the row names
row.names(wisc.data) <- wisc.df$id

# Create diagnosis vector
diagnosis <- as.numeric(wisc.df$diagnosis == "M")

# High level data exploration
dim(wisc.data)
colnames(wisc.data)
table(diagnosis)

# Check column means and standard deviations
colMeans(wisc.data)
apply(wisc.data, 2, sd)
```

```{r}
# Execute PCA, scaling if appropriate: wisc.pr
wisc.pr <- prcomp(wisc.data, center = TRUE, scale = TRUE)

# Look at summary of results
summary(wisc.pr)

```

```{r}
# Create a biplot of wisc.pr
biplot(wisc.pr)

# Scatter plot observations by components 1 and 2
plot(wisc.pr$x[, c(1, 2)], col = (diagnosis + 1), 
     xlab = "PC1", ylab = "PC2", main = "PC1 vs PC2")

# Scatter plot observations by components 1 and 3
plot(wisc.pr$x[, c(1,3)], col = (diagnosis + 1), 
     xlab = "PC1", ylab = "PC3", main = "PC1 vs PC3")

# Scatter plot observations by components 1 and 4
plot(wisc.pr$x[, c(1,4)], col = (diagnosis + 1), 
     xlab = "PC2", ylab = "PC4", , main = "PC1 vs PC4")
```

```{r}
# Set up 1 x 2 plotting grid
par(mfrow = c(1, 2))

# Calculate variability of each component
pr.var <- wisc.pr$sdev^2

# Variance explained by each principal component
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
  ylab = "Proportion of Variance Explained", 
  ylim = c(0, 1), type = "b")

# Plot cumulative proportion of variance explained
plot(cumsum(pve), xlab = "Principal Component", 
  ylab = "Cumulative Proportion of Variance Explained", 
  ylim = c(0, 1), type = "b")
```
80% of cumulative variance described by 5 principal components.

```{r}
# Scale the wisc.data data
data.scaled <- scale(wisc.data)

# Calculate the (Euclidean) distances
data.dist <- dist(data.scaled)

# Create a hierarchical clustering model
wisc.hclust <- hclust(data.dist, method = "complete")

plot(wisc.hclust)
abline(h = 20, col="red") #abline indicating 4 principal components in dendrogram
```

```{r}
# Cut tree so that it has 4 clusters: wisc.hclust.clusters
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)

# Compare cluster membership to actual diagnoses
table(wisc.hclust.clusters, diagnosis)

# Create a k-means model on wisc.data: wisc.km
wisc.km <- kmeans(scale(wisc.data), centers = 2, nstart = 20)

# Compare k-means to actual diagnoses
table(wisc.km$cluster, diagnosis)

# Compare k-means to hierarchical clustering
table(wisc.km$cluster, wisc.hclust.clusters)

```

```{r}
# Create a hierarchical clustering model: wisc.pr.hclust
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "complete")

# Cut model into 4 clusters: wisc.pr.hclust.clusters
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k = 4)

# Compare to actual diagnoses
table(wisc.pr.hclust.clusters, diagnosis)
table(wisc.hclust.clusters, diagnosis)

# Compare to k-means and hierarchical
table(wisc.km$cluster, wisc.pr.hclust.clusters)
table(wisc.km$cluster, wisc.hclust.clusters)
table(wisc.km$cluster, diagnosis)
```

